---
title: "HG002 Q100 T2T draft benchmark external evaluation variant table"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
  highlight: tango
  number_sections: no
  theme: default
  toc: yes
  toc_depth: 3
  toc_float:
    collapsed: no
    smooth_scroll: yes
---

__TODO__
- sanity checks
- summary / logging

## Loading libraries and defining inputs

```{r}
library(tidyverse)
fn_tsv_path <- snakemake@input$variant_tsv[1]
fp_tsv_path <- snakemake@input$variant_tsv[2]
fn_ref_path <- snakemake@input$ref_coords[1]
fp_ref_path <- snakemake@input$ref_coords[2]
fn_mat_path <- snakemake@input$lifted_coords[1]
fn_pat_path <- snakemake@input$lifted_coords[2]
fp_mat_path <- snakemake@input$lifted_coords[3]
fp_pat_path <- snakemake@input$lifted_coords[4]
sample_seed <- snakemake@params$seed
sample_n <- snakemake@params$sample_n
out_full_path <- snakemake@params$full
out_sampled_path <- snakemake@params$sampled
```


```{r}
make_coords_tbl <- function(anno_tsv, ref_coords, mat_coords, pat_coords){
    label <- if_else(stringr::str_detect(anno_tsv, "fn"), "fn","fp")
    ## TODO add check to make sure all fp or fn
    var_tbl <- read_tsv(anno_tsv) |> 
        rename(
            CHROM = `#CHROM`,
            INFO = `(null)`
        ) |> 
            rowwise() |> 
            mutate(INFO_md5 = digest::digest(INFO,algo = "md5"))

    grch38 <- read_tsv(ref_coords, col_names = c("GRCh38_chr", "GRCh38_start","GRCh38_end", "INFO_md5"))
    mat <- read_tsv(mat_coords, col_names = c("mat_chr", "mat_start","mat_end", "INFO_md5"))
    pat <- read_tsv(pat_coords, col_names = c("pat_chr", "pat_start","pat_end", "INFO_md5"))

    coords_tbl <- full_join(var_tbl, grch38) |> 
        full_join(mat) |> 
        full_join(pat) |> 
        mutate(label = label, SVLEN = as.integer(SVLEN))

    return(coords_tbl)
}
```

## Load and combined tables

```{r message = FALSE, warning = FALSE}
fn_coords_tbl <- make_coords_tbl(fn_tsv_path, fn_ref_path, fn_mat_path, fn_pat_path)
fp_coords_tbl <- make_coords_tbl(fp_tsv_path, fp_ref_path, fp_mat_path, fp_pat_path)
## Combining FP and FN tables 
combined_tbl <- bind_rows(fn_coords_tbl, fp_coords_tbl) |> 
        mutate(TR_anno = if_else(TRF == "1", "TR","nonTR")) |> 
        mutate(sv_type = case_when(SVTYPE == "DUP" ~ "INS",
        SVTYPE %in% c("INS","DEL") ~ SVTYPE,
        TRUE ~ "OTHER"))
```

```{r}
combined_tbl |> write_tsv(out_full_path)
```


## Random sampling

```{r}
set.seed(sample_seed)
sampled_tbl <- combined_tbl |> 
    group_by(label, TR_anno, sv_type) |> 
  slice_sample(n = sample_n)
```

```{r}
sampled_tbl |> 
    select(label, TR_anno, sv_type, CHROM, POS0, END, SVLEN, TRFdiff, TRFrepeat, TRFovl, TRFstart, TRFend, TRFperiod, TRFcopies, TRFscore, TRFentropy, INFO, GRCh38_chr, GRCh38_start, GRCh38_end, mat_chr, mat_start, mat_end, pat_chr, pat_start, pat_end) |> 
    write_tsv(out_sampled_path)
```

## Discrepancy Summary

Number of variants per strata
```{r message = FALSE, warning = FALSE}
full_counts <- combined_tbl |> 
    group_by(label, TR_anno, sv_type) |> 
    count(name = "total")

sampled_counts <- sampled_tbl |> 
    group_by(label, TR_anno, sv_type) |> 
    count(name = "sampled")

full_join(full_counts, sampled_counts)
```

SV Size distributions
```{r message = FALSE, warning = FALSE}
combined_tbl |> 
    filter(SVLEN != ".", SVTYPE %in% c("INS","DEL")) |> 
    mutate(SVLEN = as.integer(SVLEN),
    TRF = if_else(TRF == 1, "TandemRepeat","nonTR")) |> 
    ggplot() + 
    geom_vline(aes(xintercept = 300), linetype = 2) + 
    geom_vline(aes(xintercept = 5000), linetype = 2) +
    geom_histogram(aes(x = SVLEN, fill = label), bins = 100) + 
    facet_grid(SVTYPE~TRF) + 
    scale_x_log10() + 
    theme_bw()
```

```{r}
sampled_tbl |> 
    filter(SVLEN != ".", SVTYPE %in% c("INS","DEL")) |> 
    mutate(SVLEN = as.integer(SVLEN),
    TRF = if_else(TRF == 1, "TandemRepeat","nonTR")) |> 
    ggplot() + 
    geom_vline(aes(xintercept = 300), linetype = 2) + 
    geom_vline(aes(xintercept = 5000), linetype = 2) +
    geom_histogram(aes(x = SVLEN, fill = label), bins = 100) + 
    facet_grid(SVTYPE~TRF) + 
    scale_x_log10() + 
    theme_bw()
```

Size distribution of intervals for IGV
```{r message = FALSE, warning = FALSE}
combined_tbl |> 
    filter(SVLEN != ".", SVTYPE %in% c("INS","DEL")) |> 
    mutate(SVLEN = as.integer(SVLEN),
    TRF = if_else(TRF == 1, "TandemRepeat","nonTR"),
    igv_interval = GRCh38_end - GRCh38_start) |> 
    ggplot() + 
    geom_histogram(aes(x = igv_interval, fill = label), bins = 100) + 
    facet_grid(SVTYPE~TRF) + 
    scale_x_log10() + 
    theme_bw() + 
    annotation_logticks(sides = "b")
```

Sampled variants
```{r message = FALSE, warning = FALSE}
sampled_tbl |> 
    filter(SVLEN != ".", SVTYPE %in% c("INS","DEL")) |> 
    mutate(SVLEN = as.integer(SVLEN),
    TRF = if_else(TRF == 1, "TandemRepeat","nonTR"),
    igv_interval = GRCh38_end - GRCh38_start) |> 
    ggplot() + 
    geom_histogram(aes(x = igv_interval, fill = label), bins = 100) + 
    facet_grid(SVTYPE~TRF) + 
    scale_x_log10() + 
    theme_bw() + 
    annotation_logticks(sides = "b")
```

Relationship between SV size and IGV region (interval)
```{r message = FALSE, warning = FALSE}
combined_tbl |> 
    filter(SVLEN != ".", SVTYPE %in% c("INS","DEL")) |> 
    mutate(SVLEN = as.integer(SVLEN),
    TRF = if_else(TRF == 1, "TandemRepeat","nonTR"),
    igv_interval = GRCh38_end - GRCh38_start) |> 
    ggplot() + 
    geom_point(aes(x = SVLEN, y = igv_interval, fill = label),shape = 21) + 
    facet_grid(SVTYPE~TRF) + 
    scale_x_log10() + 
    scale_y_log10() + 
    theme_bw() + 
    annotation_logticks(sides = "bl")
```

## Reproducibility
TODO - add GIAB template code 

Snakemake parameters used to generate report
```{r}
snakemake
```

Saving parameters for debugging
```{r}
save.image(snakemake@params$rdata)
```
